% ------------ CONCLUSION AND FUTURE WORK ----------------
\chapter{CONCLUSION AND FUTURE WORK}%
\label{chap:conclusion_and_future_work}

% Conclusion
\section{CONCLUSION}

This project \MakeLowercase\DocumentType{} presented the design and development of \textbf{\Project{}}, an intelligent 
system aimed at addressing the limitations of traditional SMS spam filtering solutions. Unlike 
conventional binary classifiers, the proposed system focuses on multi-category SMS classification, 
enabling finer-grained differentiation between various types of spam such as phishing, 
promotional, and fraudulent messages.

The \MakeLowercase\DocumentType{} detailed a structured Software Development Life Cycle (SDLC) approach, covering 
requirement analysis, literature review, methodology, system architecture, 
implementation strategy, risk analysis, and implementation planning. The integration of multiple 
machine learning and deep learning models, combined through an ensemble-based aggregation mechanism, 
enhances robustness and classification accuracy. Furthermore, the 
incorporation of Explainable Artificial Intelligence (XAI) techniques improves system transparency 
and user trust by providing interpretable explanations for model decisions.

Overall, the proposed system is technically feasible, ethically responsible, and aligned with 
modern software engineering and machine learning best practices.

% Future Work
\section{FUTURE WORK}

Although the proposed system fulfills its stated objectives, several extensions and improvements 
can be explored in future work:
\begin{itemize}
    \item \textbf{Multilingual SMS Detection:} Extending the system to support multiple languages to enhance real-world applicability.
    \item \textbf{Transformer-Based Models:} Incorporating advanced deep learning architectures such as BERT or other transformer models for improved contextual understanding.
    \item \textbf{Real-Time Deployment:} Integrating the system with mobile network operators or messaging platforms for real-time SMS filtering.
    \item \textbf{Adaptive Learning:} Implementing online learning mechanisms to handle evolving spam patterns and concept drift.
    \item \textbf{Enhanced Explainability Interfaces:} Developing richer visualization dashboards for administrators and end users.
\end{itemize}
\enumitemEndSpacing{}

These enhancements would further strengthen the effectiveness, scalability, and usability of the SMS Spam Shield system.

% Final Remarks
\section{FINAL REMARKS}

The proposed SMS Spam Shield system demonstrates how machine learning, ensemble methods, and 
explainable AI can be combined to build transparent and reliable security-oriented applications. 
The project provides a strong foundation for future academic research and practical deployment 
in SMS security systems.
