\chapter{TECHNICAL DETAILS}%
\label{chap:technical_details}

% Overview
\section{OVERVIEW}

This chapter provides detailed technical specifications for the \textbf{\Project{}} system. 
It covers the dataset structure, preprocessing methodology, feature extraction techniques, 
and classification pipeline architecture that form the foundation of the spam detection model.

% Dataset Structure
\section{DATASET STRUCUTURE}

The SMS dataset used in this project follows a structured tabular format. Each record represents 
a single SMS message with an associated category label.

\begin{table}
\centering
\begin{tabular}{|c|p{5cm}|p{5cm}|}
\hline
\textbf{Field Name} & \textbf{Description} & \textbf{Data Type} \\
\hline
sms\_id & Unique identifier for SMS & Integer \\
\hline
message\_text & Raw SMS content & String \\
\hline
category & SMS classification label & Categorical \\
\hline
\end{tabular}%
\caption{SMS Dataset Schema}
\label{table:dataset_schema}
\end{table}

% SMS Category Definations
\section{SMS CATEGORY DEFINATIONS}

The system supports multi-category classification. The primary categories used are:
\begin{itemize}
    \item \textbf{Legitimate:} Personal or informational messages
    \item \textbf{Promotional:} Marketing or advertising messages
    \item \textbf{Phishing:} Messages attempting to steal sensitive information
    \item \textbf{Scam/Fraud:} Messages involving financial deception
    \item \textbf{Transactional:} One-time password (OTP\nomenclature{OTP}{One Time Password}) or service notifications
\end{itemize}
\enumitemEndSpacing{}

% Preprocessing Pipeline Pseudocode
\section{PREPROCESSING PIPELINE PSEUDOCODE}

\begin{verbatim}
    Input: Raw SMS Text
    Output: Cleaned Token List

    1. Convert text to lowercase
    2. Remove punctuation and special characters
    3. Tokenize text into words
    4. Remove stop-words
    5. Apply lemmatization
    6. Return cleaned tokens
\end{verbatim}

% Feature Vector Representation
\section{FEATURE VECTOR REPRESENTATION}

\textbf{TF-IDF Vector Example:}
\begin{itemize}
    \item SMS\texttt{:} ``Win a free prize now''
    \item Extracted Features: [win, free, prize, now]
    \item TF-IDF Vector: Sparse numerical representation
\end{itemize}
\enumitemEndSpacing{}

\textbf{Sequence Encoding Example:}
\begin{itemize}
    \item Tokenized SMS\texttt{:} [win, free, prize, now]
    \item Encoded Sequence: [45, 112, 78, 203]
    \item Padded Sequence Length: 50
\end{itemize}
\enumitemEndSpacing{}

% Sample Input and Output
\section{SAMPLE INPUT AND OUTPUT}

\textbf{Sample Input SMS:}

``Congratulations! You have won a free gift card. Click the link to claim now.''

\textbf{System Output:}
\begin{itemize}
    \item Predicted Category: Phishing
    \item Confidence Score: 0.94
    \item Important Tokens: win, free, click, claim
\end{itemize}
\enumitemEndSpacing{}

% Additional Notes
\section{ADDITIONAL NOTES}

All sample data shown in this appendix is anonymized and used solely for academic and experimental 
purposes. No personally identifiable information is stored or processed by the system.
